The persuasion of symbolic elements in government public communications, and their moderation by personal relevance

### 2018-07-16 paper3_relevance and symbols_01.Rmd
Add files via upload

---
title: "The persuasion of symbolic elements in government public communications, and their moderation by personal relevance"
author: '**Saar Alon-Barkat**, PhD candidate; the Hebrew University of Jerusalem;   saar.barkat@mail.huji.ac.il'
date: " "
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: yes
bibliography: phd_paper_2.bib
#csl: apsa.csl
urlcolor: blue
---

@@

Last edited at `r Sys.Date()`.[^footnote_scripts]

[^footnote_scripts]: Original R scripts available at:  https://github.com/saaralonbarkat/symbols_and_relevance

@@

*Recent studies have demonstrated that symbolic elements in government public communications can enhance citizens' trust in government organizations and their policies. They also suggested that the emotive effect of these symbols can compensate for organizations' poor performance and inefficient policies, thereby eliciting undue trust. This study explores the boundaries of governments' persuasion of citizens through symbolic communication. Building on social psychology literature, I hypothesize that citizens are less susceptible to symbols and more likely to scrutinize the substantive content of the communication, the more they perceive the policy issue in the communication as personally relevant. I test this expectation through a survey experiment, where I examine Israeli citizens' trust in the Environment Protection Ministry's policy plans for reducing air pollution in a highly polluted area. The perceived personal relevance of the policy is operationalized both experimentally, and by comparing between citizens residing in the polluted area and elsewhere. The results indicate that even high levels of perceived personal relevance do not attenuate the positive effect of symbols. These findings may suggest that the boundaries of the effect of symbolic communication are wider than expected, which carries significant implications.*

@@

A growing public administration literature explores whether government public communications can enhance citizens' trust in public organizations and policies. Whereas most of the priovious studies have focused on the *informative* aspects of government communications [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @grimmelikhuijsen_2015; @ho_2017; @james_2014; @james_2011], recent studies further pointed to the emotive effect of *symbolic* elements entangled in them, such as brand names [@teodoro_2018], logos [@karens_2016], and figures [@alonbarkat_2017]. Such symbols and others exist in almost every government communication, and most prominently in public campaigns, which have been increasingly used by public-sector organizations [@eshuis_2012; @marland_2017].

Persuasion through symbolic communications and branding can be perceived favorably, insofar as it enables governments to engage more effectively with relevant publics, thereby to increase their trust and enhance their collaboration. The effectiveness of symbols, as opposed to communication of "objective" information, stems from their potency to overcome citizens' negative biases. On the other hand, employing a broader perspective of democratic theory, one can argue that in the public-sector domain, the emotive effect of symbols undermines the fundamental normative expectation that citizens' evaluation of their government should be based on their critical judgment. Furthermore, previous studies have demonstrated that the positive effect of symbols can compensate for organizations' poor performance (@alonbarkat_2017) and for logically unpersuasive informative content (@alonbarkat_2018). Accordingly, symbolic communication can create a misalignment between citizens' positive perceptions and the "real world", and mitigate justified public criticism of government failures.

@@

The Elaboration Likelihood Model (ELM), developed in the late 1970s by social psychology scholars Richard Petty and John Cacioppo, provides a theoretical framework for investigating the effects of communications on people's attitudes [@petty_1986; @petty_2011]. According to ELM, people may respond differently to a communication, depending on the amount of thinking that they dedicate to it. When people have high motivation and ability to elaborate on a message, they are likely to focus on the arguments in the communication that are central to the merits of the issue, and scrutinize them ("central route"). Conversely, when they are unmotivated or unable to elaborate, they may rely on available heuristic cues that require relatively little amount of cognitive effort ("peripheral route"). These two routes for persuasion are regarded as two ends of an "elaboration continuum". People may also be located between these two edges, and then a mixture of these processes will determine their persuasion.

@@

@@

<br>
**Figure 2: Experimental design for policy plan communications**
![](D:/UNIVERSITY/R/SVIVA/experiment 2 01-2018/myfigures/experiment_flow_relevance.png)

@@

@@

@@

Public communications has been proposed as a remedy for citizens’ low trust in government, and their underestimation of public-service performance. Congruently, governments' investments in public communications, and specifically in branding and advertising, has increased over the past generation in many economically-developed democratic countries. This study contributes to the growing public administration research on citizens' responses to government public communications, and specifically to symbolic elements entangled in these communications. I explored the expectation, derived from social psychology, that citizens are less affected by symbols and more likely to scrutinize the content of the communication the more they perceive it as relevant to their own lives. I put these hypotheses to rigorous empirical testing, using a randomized survey experiment focusing on the Israeli Environmental Protection Ministry. I exploited the natural variation in citizens' perceptions of the personal relevance of the ministry’s policy plan regarding air pollution in a specific area. In addition to utilizing these natural differences, I also manipulated the personal relevance of the policy plans. To the best of my knowledge, this is the first study to assess the moderating role of perceived personal relevance on citizens' responses to government symbolic communications or branding.

@@

@@

>>>>

---
title: "The persuasion of symbolic elements in government public communications, and their moderation by personal relevance"
author: '**Saar Alon-Barkat**, PhD candidate; the Hebrew University of Jerusalem;   saar.barkat@mail.huji.ac.il'
date: " "
output:
  word_document: default
  pdf_document:
    toc: yes
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
link-citations: yes
bibliography: phd_paper_2.bib
#csl: apsa.csl
urlcolor: blue
---

@@

Draft, last edited at `r Sys.Date()`.

@@

*Recent studies demonstrated that symbolic elements in government communications can enhance citizens' trust in government organizations and policies. The emotive effect of symbols can further compensate for organizations' poor functioning, thereby eliciting undue trust. This study explores the boundaries of governments' persuasion of citizens through symbolic communication. Building on social psychology, I hypothesize that citizens are less susceptible to symbols, the more they perceive the policy issue in the communication as personally relevant. I test this expectation through a survey experiment, where I examine Israeli citizens' trust in the Environment Protection Ministry's policy plans for reducing air pollution in a highly polluted area. The policy's perceived personal relevance is operationalized both experimentally, and by comparing between citizens residing in the polluted area and elsewhere. The results indicate that even high levels of perceived personal relevance do not attenuate the symbols' effect, which entails that the boundaries of persuasion through symbols are wider than expected.*

*Keywords: Government communications; Government branding; Trust; Experiment*

@@

A growing public administration literature explores whether government public communications can enhance citizens' trust in public organizations and policies. Whereas most of the studies have focused on the *informative* aspects of government communications [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @porumbescu_2017a; @porumbescu_2017b; @porumbescu_2018; @grimmelikhuijsen_2015; @ho_2017; @james_2014; @james_2011], recent studies further pointed to the emotive effect of *symbolic* elements entangled in them, such as brand names [@teodoro_2018], logos [@karens_2016], and figures [@alonbarkat_2017]. Such symbols and others exist in almost every government communication, and most prominently in public campaigns and branding activities,[^footnote_branding] which have been increasingly used by public-sector organizations [@eshuis_2012; @marland_2017].

Governments' employment of symbolic communications and branding can be perceived favorably, insofar as it enables organizations to engage more effectively with citizens and overcome their negative prior beliefs and biases against the public sector [@marvel_2015a; @marvel_2015; @hvidman_2016]. Accordingly, it can mitigate citizens' distrust in government organizations and enhance their collaboration with their policies. On the other hand, one can argue that the persuasion of citizens through symbols is problematic, since the emotive effect of symbols undermines the fundamental normative expectation that citizens' evaluation of their government should be based on their critical judgment. Furthermore, symbols can compensate for organizations' poor performance and for logically unpersuasive informative content, as empirically demonstrated in previous studies [@alonbarkat_2017; @alonbarkat_2018]. They can yield a misalignment between citizens' positive perceptions and the "real world". This raises the concern that government organizations will strategically use symbols to manipulate public opinion and mitigate justified public criticism of their failures.

[^footnote_branding]: A symbol can be as *"any object used by human beings to index meanings that are not inherent in, nor discernible from, the object itself"* [@elder_1983, pp. 28-29]. Eshuis and Klijn [-@eshuis_2012, p. 19] define brand as *a symbolic construct that consists of a name, term, sign, symbol, or design, or combination of these, created deliberately to identify and phenomenon and differentiate it from similar phenomena by adding particular meaning to it*. Brand elements, therefore, can be regarded as symbolic elements.

@@

The Elaboration Likelihood Model (ELM), developed in the late 1970s by social psychology scholars Richard Petty and John Cacioppo, provides a comprehensive theoretical framework for investigating the effects of communications on people's attitudes [@petty_1986; @petty_2011]. According to ELM, people may respond differently to a communication, depending on the amount of thinking that they dedicate to it. When people have high motivation and ability to elaborate on a message, they are likely to focus on the arguments in the communication that are central to the merits of the issue, and scrutinize them ("central route"). Conversely, when they are unmotivated or unable to elaborate, they may rely on available heuristic cues that require relatively little amount of cognitive effort ("peripheral route"). These two routes for persuasion are regarded as two ends of an "elaboration continuum". People may also be located between these two edges, and then a mixture of these processes will determine their persuasion.

@@

[Figure 1 about here]

@@

[Figure 2 about here]

@@

[Table 1 about here]

[Figure 3 about here]

@@

[Table 2 about here]

@@

Public communications has been proposed as a remedy for citizens’ low trust in government, and their underestimation of public-service performance. Congruently, governments' investments in public communications, and specifically in branding and advertising, has increased over the past generation in many economically-developed democratic countries. This study contributes to the growing Public-Administration research on citizens' responses to government public communications, and specifically to symbolic elements entangled in these communications. I explored the expectation, derived from social psychology, that citizens are less affected by symbols and more likely to scrutinize the content of the communication the more they perceive it as relevant to their own lives. I put these hypotheses to rigorous empirical testing, using a randomized survey experiment focusing on the Israeli Environmental Protection Ministry. I exploited the natural variation in citizens' perceptions of the personal relevance of the ministry’s policy plan regarding air pollution in a specific area. In addition to utilizing these natural differences, I also manipulated the personal relevance of the policy plans. To the best of my knowledge, this is the first study to assess the moderating role of perceived personal relevance on citizens' responses to government symbolic communications or branding.

@@

The results of this study, if generalizable, have significant implications. They challenge the rationalistic view of government direct communications with citizens as a means to inform citizens about government organizations actions and promote transparency and accountability. Instead, they suggest that more often than not, citizens rely on brands and other symbolic elements in public communications, even with regard to policy issues that have direct implications on their own lives. Altogether, this study highlights the role played by government branding and other forms of symbolic communication in shaping citizens view of governments. It suggests that the boundaries of persuasion via symbols and branding are much wider than expected, based on previous social psychology and marketing studies. From the practical point of view of government organizations, this entails that organizations can overcome citizens distrust by investing in symbolic communications. Nonetheless, employing a broader perspective of democratic theory, these conclusions are disturbing insofar as they buttress the concern that government organizations will deliberately use symbolic communications to cause citizens to view them more favorably regardless (or even in contrast) to their actual capacities and achievements, namely as "propaganda". For good or for ill, citizens are affected by the emotional connections and associations of symbols.

@@

#Figures and tables

**Figure 2: Experimental design for policy plan communications**
![](D:/UNIVERSITY/R/SVIVA/experiment 2 01-2018/myfigures/experiment_flow_relevance.png)
